---
title: "Moving Averages in R"
author: "William Chiu"
date: "5/7/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stargazer)
```

## Data and Noise

Suppose $y$ is a linear function of $x_1$ and random error:

$$y = \beta_0 + \beta_1 x_1 + \epsilon$$
```{r}
set.seed(1)

nobs <- 1000

x1 <- log(1:nobs) + rnorm(nobs, 0, 0.25)
y <- 2 + 3 * x1 + rnorm(nobs)

df <- data.frame(t=1:nobs, y=y, x1=x1)

rm(x1, y)
```

But also suppose that $x_1$ is not directly observable, instead we observe its noiser cousin that contains random errors. For example, the instrument that measures $x_1$ suffers from random imprecision.

```{r}
df$x1_noisy <- df$x1 + rnorm(nobs,0)
```

Since $x_1$ is observed with noisy imprecision, this degrades the OLS fit.

```{r}
true_model <- lm(y ~ x1, data=df)

noisy_model <- lm(y ~ x1_noisy, data=df)

```

```{r echo=FALSE, results='asis'}
stargazer(true_model, noisy_model,
          title="True and Noisy Predictors",
          type = "latex",
          float = FALSE,
          report = "vcs*",
          no.space = TRUE,
          header=FALSE,
          single.row = FALSE,
          font.size = "small",
          intercept.bottom = F,
          digits = 3, omit.stat=c("f")
)
```

## Visualization

Plot $x_1$ and $x_1^{noisy}$ over time.

```{r}
df_long <- pivot_longer(df, -t) %>% filter(name %in% c('x1','x1_noisy'))

ggplot(df_long, aes(x=t, y=value, group=name, color=name)) +
  geom_line(aes(linetype=name)) +
  scale_linetype_manual(values=c("solid", "dashed"))+
  scale_color_manual(values=c('black', 'green')) +
  theme_bw()
```



## Filter and Moving Averages

We could apply a filter on $x_1^{noisy}$ to remove some of the "jumpiness". First, we apply a backward-looking moving average with a k-period window:

```{r}
k_param <- 12

df$x1_noisy_ma <- stats::filter(x=ts(df$x1_noisy),
                                 filter=rep(1/k_param, k_param),
                                 method="convolution",
                                 sides=1)

knitr::kable(head(df, k_param+5))
```

Second, we apply a "centered" moving average that includes both past and future periods.


```{r}
df$x1_noisy_ma_ctr <- stats::filter(x=ts(df$x1_noisy),
                                 filter=rep(1/k_param, k_param),
                                 method="convolution",
                                 sides=2)

knitr::kable(head(df, k_param+5))
```

The filters created `NA` values that should be removed before refitting the models.

```{r}
df_filtered <- df %>% drop_na()

knitr::kable(head(df_filtered))
```



## More Visualization

Plot $x_1$, $x_1^{noisy}$, and the filtered predictors  over time.

```{r}
df_long2 <- pivot_longer(df_filtered, -t) %>%
  filter(name %in%  c('x1','x1_noisy', 'x1_noisy_ma', 'x1_noisy_ma_ctr'))

ggplot(df_long2, aes(x=t, y=value, group=name, color=name)) +
  geom_line(aes(linetype=name)) +
  scale_linetype_manual(values=c("solid", "dashed", "solid", "solid"))+
  scale_color_manual(values=c('black', 'green', 'purple', 'magenta')) +
  theme_bw()
```

## Refit models with MA predictors

```{r}
ma_model <- lm(y ~ x1_noisy_ma, data=df_filtered)

ma_ctr_model <- lm(y ~ x1_noisy_ma_ctr, data=df_filtered)

```

```{r echo=FALSE, results='asis'}
stargazer(true_model, noisy_model, ma_model, ma_ctr_model,
          title="True, Noisy, and Filtered Predictors",
          type = "latex",
          float = FALSE,
          report = "vcs*",
          no.space = TRUE,
          header=FALSE,
          single.row = FALSE,
          font.size = "small",
          intercept.bottom = F,
          digits = 3, omit.stat=c("f")

)
```

## Optimal Filter via 5-fold CV MSE

The hyper-parameter `k_param` controls the smoothness of the `ma` predictor. We could choose the optimal `k_param` by trying different values and measuring the 5-fold cross-validation error for each value. The code below finds the optimal `k_param` for the backward-looking moving average.

```{r}
cv_5_fold <- function(dataframe, ma_window){

  dataframe[,'x1_noisy_ma'] <- stats::filter(x=ts(dataframe[,'x1_noisy']),
                                 filter=rep(1/ma_window, ma_window),
                                 method="convolution",
                                 sides=1)  
  
  train <- dataframe %>% drop_na()

  mod <- glm(y ~ x1_noisy_ma, data=train)
  
  set.seed(123)

  cv_mod <- boot::cv.glm(train, mod, K=5)

  return(data.frame(ma_window=ma_window, MSE=cv_mod$delta[1]))
}
```

Let's try the moving average windows from 2 to 120 (in increments of 2):

```{r}
tuning_ma_window <- lapply(seq(2,120,2), function(x){
  cv_5_fold(dataframe=df, ma_window=x)
})

tuning_ma_window_df <- bind_rows(tuning_ma_window)
```

Plot the window against 5-fold CV MSE:

```{r}
ggplot(tuning_ma_window_df, aes(x=ma_window,y=MSE)) +
  geom_line() +
  ylab("5-fold CV MSE") +
  theme_bw()
```

There does not appear to be a meaningful change in CV MSE after `k=24`. Hence, we choose 24 as the optimal moving average window.

```{r}
k_param <- 24

df$x1_noisy_ma_cv <- stats::filter(x=ts(df$x1_noisy),
                                 filter=rep(1/k_param, k_param),
                                 method="convolution",
                                 sides=1)

df_filtered_cv <- df %>% drop_na()

df_long_cv <- pivot_longer(df_filtered_cv, -t) %>%
  filter(name %in%  c('x1','x1_noisy', 'x1_noisy_ma', 'x1_noisy_ma_cv'))

ggplot(df_long_cv, aes(x=t, y=value, group=name, color=name)) +
  geom_line(aes(linetype=name)) +
  scale_linetype_manual(values=c("solid", "dashed", "solid", "solid"))+
  scale_color_manual(values=c('black', 'green', 'purple', 'magenta')) +
  theme_bw()

ma_model_cv <- lm(y ~ x1_noisy_ma_cv, data=df_filtered_cv)


```

```{r echo=FALSE, results='asis'}
stargazer(true_model, noisy_model, ma_model, ma_model_cv,
          title="True, Noisy, and Filtered Predictors",
          type = "latex",
          float = FALSE,
          report = "vcs*",
          no.space = TRUE,
          header=FALSE,
          single.row = FALSE,
          font.size = "small",
          intercept.bottom = F,
          digits = 3, omit.stat=c("f")

)
```


